Characterizing Stellar Streams by ML

### 2023-09-06 src/tex/ms.tex
[showyourwork] overleaf sync

% Change the section names
\renewcommand{\sectionautorefname}{\S\!}%
\renewcommand{\subsectionautorefname}{\S\!\!}%
\renewcommand{\subsubsectionautorefname}{\S\!}%
\renewcommand{\equationautorefname}{Eq.}
\renewcommand{\tableautorefname}{Tab.}

@@

    \begin{quotation}

        A stellar stream is the association of stars tidally stripped from a
        common progenitor system that orbits in an underlying galactic
        gravitational field \citep{Johnston1998, HelmiWhite1999}. Stream
        progenitors are gravitationally bound systems, such as satellite dwarf
        galaxies or star clusters \citep{Odenkirchen+2001, Majewski+2003}. In
        the absence of external interactions these systems are stable, though
        still capable of mass loss driven by internal processes
        \citep{Hills1975, HeggieHut2003}, e.g. relaxation and binary
        interactions. Considering external interactions, these progenitor
        systems may be disrupted: completely, as in the merger events that
        formed the Galactic stellar halo \citep{Lynden-Bell1967,
        SearleZinn1978}; or partially, such as tidal disruption forming a
        stellar stream.

        In the approximation of a smooth galactic gravitational field, the
        progenitor's extent is limited to the tidal radius. Through processes
        internal and external, progenitor stars can achieve sufficient energy to
        bring them beyond the tidal boundary, thus ``escaping'' into the
        galactic field. Escape primarily occurs at both an inner and outer
        saddle point of the combined potential -- e.g. the L2 and L3 Lagrange
        points for a spherical galactic field \citep{Ross+1997,
        FukushigeHeggie2000, BinneyTremaine2008}. Generally, the initial phase
        space position of the escaped star is very similar to that of the
        progenitor: the configuration position is the tidal boundary and the
        star has a low peculiar velocity. Since both the escaped star and its
        progenitor orbit in the same underlying galactic potential and have
        similar phase space positions their orbits are similar
        \citep{BinneyTremaine2008}. Given sufficient time the phase space
        separation will increase and an escaped star will lead or trail the
        position of the progenitor, depending on the location from which it
        escaped. Therefore, a collection of tidally stripped stars will form
        extended structures, aka tails or streams, that are few to thousands of
        parsecs in length \citep{Johnston1998, HelmiWhite1999, Bovy2014}.

        To be a coherent, observable structure the peculiar velocity dispersion
        of escaped stars relative to the progenitor must be relatively low, but
        does vary from system to system. Stars in kinematically hot streams
        (e.g. from progenitors with larger kinematic dispersions) have more
        dissimilar orbits from each other and the progenitor than do stars in
        cold streams, with smaller dispersions \citep{Johnston1998,
        HendelJohnston2015}.  The orbit of a star depends on its phase space
        position and the gravitational potential; therefore measuring a star's
        position and velocity and observing its orbit reveals the local
        potential, which is normally dominated by the Galaxy (primarily the dark
        matter halo at large distances) \citep{Gibbons+2014}. However, an orbit
        is impractical to observe on human timescales. Cold streams offer a
        workaround.  Like measuring one star at many points in one orbit over an
        extended period, streams are many stars measured at different points
        along approximately one orbit at the same time. Therefore stellar
        streams are one of the most promising means to study the Galactic
        gravitational potential: the stellar components as well as the dark
        matter \citep{Johnston2016}.

        On the observational front there has been a large concerted effort in
        the field to find new stellar streams \citep[for an extensive list of
        known Galactic streams see][]{Mateu2022}, extend the detected extents of
        known streams \citep[e.g. for \stream{Palomar 5} alone:][]{Rockosi+2002,
        Ibata+2017, Grillmair+2006, Carlberg+2012, Starkman+2019}, and increase
        the general purity of stream catalogs.

    \end{quotation}

    \todo[inline]{--- Rephrase}

    \paragraph{one orbit and a single population}

        Stellar streams consist of stars that are tidally disrupted from a
        progenitor system.  To a first approximation, the stars occupy a single
        orbit, that of their progenitor.  Additionally, as stars formed in the
        progenitor they retain its photometric and chemical characteristics:
        which for globular clusters means following an isochrone.  In this
        simplistic picture, astrometrically characterizing a stream requires
        only finding it's orbit. Member stars fall directly on the orbit and the
        isochrone, making membership likelihood determination trivial.

        % Observational errors complicate this picture. Instead of a certain
        % membership calculation, we must build a probabilistic model, assigning
        % a likelihood that an observed star falls along the stream's orbit.

    \paragraph{similar orbits and a somewhat noisy population}

        Occupying a single orbit implies that streams remain a point in action
        space, and having a common origin would thus not become extended
        structure. This is not true.  Differences in stream stars' orbits from
        the progenitor's arise from various factors including internal dynamics
        of the progenitor and the host Galactic potential.

        An intrinsic width, in configuration and kinematic space.

        Centered on a track, not an orbit.

        This defines an astrometric phase-space hyper-tube \cite{STREAMFINDER},
        with a distribution radial to the track.

        Instead of characterizing the stream by its orbit, we characterize it by
        it's track and the distribution around that track.  Computing membership
        likelihoods now requires fitting the track and distribution. This cannot
        be done without then also fitting the field of stars in which the stream
        lives, since the stream and field cannot be clearly separated.

    \paragraph{gaps orbits and a somewhat noisy population}

        This is still a simplistic picture of streams.

        Perturbations to the Galactic potential, in the form of time evolution
        and non-axisymmetry, as well as substructure can disrupt the stellar
        stream.  Even most perturbations don't ruin the idea of the phase-space
        hypertube. They either drop the mixture weight (rephrase, not yet
        introduced) to 0 or change the width of the hypertube.

    \paragraph{Review of detection methods}

    Review of available data:
    \begin{enumerate}
        \item Gaia. Streams in Gaia have on-sky positions and proper motions,
        and parallaxes, though for most streams the parallax errors are
        prohibitively large. Likewise, most streams in Gaia do not have good
        radial velocity measurements. The photometry is generally poor,
        generally necessitating cross-matching with other datasets.
        \item S5. This is a much smaller dataset. It has deeper photometry and
        radial velocity measurements.
    \end{enumerate}

@@

        Per \autoref{eq:stream_and_bkg_prob} the stream is an independent
        addition of both astrometric and photometric models.  In astrometrics
        the stream is a Gaussian track in $\phi_2$ and parallax $\varpi$, using
        the model defined in \autoref{ssub:method:astrometric_model:on_stream}.
        We use a 4-layer \texttt{LinTanh} with shape (1 input, 64 hidden, 4
        output), where the 4 outputs are $[\mu_{\phi_2}, \sigma_{\phi_2},
        \mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.  The network is small, but
        sufficiently large to permit 15\% dropout and retain robust predictions.
        The photometric model is a distance modulus-shifted isochrone, using the
        model outlined in  \autoref{ssub:method:photometric_model:on_stream}. We
        use the same isochrone parameters as were used to simulate the stream.
        For this mock stream we demonstrate simultaneous astrometric and
        photometric modeling, linking the distance modulus and parallax track,
        as explained in \autoref{ssub:method:linking_to_astrometrics}. Thus the
        photometric model does not have its own network, using instead the
        outputs $[\mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.

        The total background model has two pieces, an astrometric and a
        photometric model.  As a pre-training phase the photometric normalizing
        flow model was fitted to an off-stream selection and is now held fixed.
        The astrometric model is split in two components, the first is a uniform
        distribution in $\phi_2$, and the second is an exponential distribution
        for $\varpi$.  The uniform distribution does not have any parameters,
        and thus no neural network.  The exponential distribution has only the
        slope parameter and we use a 3-layer \texttt{LinTanh} with shape (1, 32,
        1) and 15\% dropout. Background and stream are combined in a mixture
        model whose mixture parameters are a 4-layer (1, 64, 1) \texttt{LinTanh}
        network. We train the mixture model with an \texttt{AdamW} optimizer
        with learning rate of $5\times10^{-3}$ and a scheduler to periodically
        and temporarily increase the learning rate, encouraging the optimizer
        towards better minima.

@@

        \begin{figure*}
            \script{mock/plot/results.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/mock/results.pdf}
            \caption{
                %
                \textbf{Top panel: }%
                    The stream mixture coefficient $f(\phi_1)$ (black dotted
                    line) predicted by the model, overplotted on the true stream
                    fraction (black-outlined histogram), calculated as the ratio
                    of stream stars to total stars in 55 bins over the $\phi_1$
                    range of the isochrone.  \textbf{Middle panel: }%
                    $\phi_2(\phi_1)$ over the full range of $\phi_1$. The data
                    color and transparency is set by the model's MLE stream
                    probability (with dropout turned off). Plotted behind the
                    data are the true labels for the stream stars (large black
                    circles).  The true labels exactly limn the model-identified
                    stream stars.  The MLE stream track (black dashed line) is
                    over-plotted on the probability-colored data.
                %
                \textbf{Lower panel 3: }%
                    $\varpi_2(\phi_1)$ over the full range of $\phi_1$. Like for
                    $\phi_2$, the data color and transparency is set by the MLE
                    probability, with the true labels (large black circles)
                    under-plotted and MLE track (black dashed line)
                    over-plotted.
                %
                \textbf{Upper Histograms: }%
                    $\phi_2$ histograms for each model component (blue
                    background, red stream), in 4 $\phi_1$ ranges across the
                    data set.  The height of each bin is the
                    probability-weighted density.  The background is a uniform
                    distribution, while the stream is a Gaussian. In projection
                    across the $\phi_1$ range it can appear multi-modal (panel
                    2) if there are large gaps or `smeared' out (panel 3) if the
                    Gaussian's mean changes significantly over the range.
                %
                \textbf{Upper Middle Histograms: }%
                    Parallax $\varpi$ histograms for each model component (blue
                    background, red stream), in 4 $\phi_1$ ranges across the
                    data set.  The height of each bin is the
                    probability-weighted density.
                %
                \textbf{Lower Middle and Lower Histograms:}%
                    Photometric coordinate $(g, r)$ plots for each model
                    component in 4 $\phi_1$ ranges across the data set. The top
                    row plots the stream stars (red) over the background. The
                    lower row plots the background stars (blue) over the stream
                    stars to show that the how the two distributions mix.
                %
            }
            \label{fig:mock_data_result}
            TODO:
            \begin{enumerate}
                \item the linear density in the stream-fraction plot
            \end{enumerate}
        \end{figure*}

@@

    Having successful applied the density modeling framework to a mock stream,
    we now apply the method to the stellar stream GD-1.
    % TODO: some reiteration on the importance of GD-1
    We discuss the data selection in \autoref{sub:results_gd1:data}, model
    specifics in \autoref{sub:results_gd1:model}, and the results in
    \autoref{sub:results_gd1:results}.

@@

        We utilize data from \Gaia{} DR3 for the astrometric coordinates of each
        star.  To obtain more accurate color information, we cross-match the
        \Gaia{} field on \PanStarrs{} \citep{PanSTARRS1} using \Gaia's provided
        \texttt{Best-neighbors} catalog.  To limit the field to a region known
        to contain \stream{GD-1} we adopt broad cuts on the cross-matched data.
        We perform these rough cuts by requiring the data lie within the
        phase-space cube:
        \begin{itemize}
            \item $\phi_1 \in [-100, 40] \ \rm{\degree}$,
            \item $\phi_2 \in [-9, 5] \rm{\degree}$,
            \item $\varpi \in [-10, 1] \ \rm{mas}$,
            \item $\rm{bp-rp} \in [-1, 3] \ \rm{mag}$
            \item $g, r \in [0, 50] \ \rm{mag}$, and not \texttt{null}.
        \end{itemize}
        where we use \package{gala}'s \citep{gala, galav1.3}
        \code{GD1Koposov10} frame \citep{Koposov+2010} to define $\phi_1$,
        $\phi_2$, and associated proper motions.  For downloading purposes we
        split the query into 10 sub-regions in even $\phi_1$ increments. A plot
        of the $\phi_1, \phi_2$ regions is included in the
        \href{\GitHubURL{}}{\package{ShowYourWork}} repository.  Stream-specific
        coordinates ($\phi_1, \phi_2$) are not included in \Gaia.  One could
        embed the coordinate transform in the ADQL \citep{ADQL2.0}, or define
        great-circle arcs in ICRS \citep{ICRS1997}. Since only rough cuts are
        required and since the sub-region boxes are sufficiently small for a
        reasonable flat-sky approximation  we simply translate the ($\phi_1,
        \phi_2$) boxes to ICRS.  We correct the \PanStarrs{} photometry for dust
        extinction using the \textit{Bayestar19} \citep{Green+2019} models in
        the \package{dustmaps} package \citep{Green2018}.  \textit{Bayestar19}
        is a three-dimensional dust map of the Milky Way for which we use
        \Gaia's parallax distance information.  We include the $1-\sigma$ errors
        as a source of uncertainty in $\mbs{\Sigma}_n$, the per-star photometric
        uncertainty.
        % TODO: actually implement that in the model
        The result is a field populated by
        \variable{output/gd1/ndata_variable.txt} stars, from which we would like
        to identify stream members.

        We create data masks for a more refined selection around \stream{GD-1}.
        The two mask types, proper motion and photometric, are show in
        \autoref{fig:gd1-data_selection}.  The proper motion masks select
        rectangular regions, containing the primary overdensity of \stream{GD-1}
        and a surrounding area following the tail of the distribution.  The
        tighter mask (red) has corners $(\mu_{\phi_1}, \mu_{\phi_2}) =
        [(-15,-10) (-4.5, -2)] \ \unit{mas/yr}$, while the larger mask (yellow)
        has $(\mu_{\phi_1}, \mu_{\phi_2}) = [(-15,-5) (-5, -1)] \
        \unit{mas/yr}$.  In photometric coordinates we define a mask around
        % TODO: make these variables, like in mock
        a 12 Gyr, [Fe/H] = -1.3 isochrone at a distance of 7.8 kpc. The
        selection region is a 0.2 dex orthogonal buffer around the MIST
        isochrone \citep[using]{brutus}, which is illustrated in the
        2nd-from-the-top-left plot in \autoref{fig:gd1-data_selection}.  In
        addition, to avoid detailed modeling of the completeness as a term in
        $\prior_{obs}$ (from \autoref{isochrone_prior_components}) we mask out
        photometry with $G_{\Gaia{}} > 20$.  With combinations of these masks
        the stream has much larger signal-to-noise, though is still only a small
        percent of the total data.

        \begin{figure*}[h]
            \script{gd1/plot/data_selection.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/gd1/data_selection.pdf}
            \caption{%
                Data selections for \stream{GD-1}, within the data cube
                described in \autoref{sub:results_gd1:data}.
                %
                \textbf{Row 1 Panels:}
                Proper motion selection boxes in red and yellow.  The right plot
                shows the application of these selections.  \stream{GD-1} is
                apparent from around $-60$ to $-10$ deg in the red selection and
                present, if less clearly, in the yellow selection.
                %
                \textbf{Row 2 Panels:}
                Photometric selection (blue curve). The selection is based on a
                12 Gyr, [Fe/H] = -1.3 MIST isochrone, buffered by 0.2 dex.  The
                right plot shows the application of this selection. By eye, the
                stream is a small overdensity extending from $-50$ to $0$ deg.
                %
                \textbf{Row 3 Panels:}
                Applying the combination of red astrometric and blue photometric
                selections, plotted for astrometric coordinates $\phi_2, \varpi,
                \mu_{\phi_1}, \mu_{\phi_2}$.  The stream is an identifiable
                overdensity in all phase-space coordinates except $\varpi$ where
                the errors are significant.
                %
                \textbf{Row 4 Panels:}
                Applying the yellow astrometric and blue photometric selections,
                plotted for the same astrometric coordinates as the above row.
                In these coordinates the stream is clearly observable in
                $\phi_2$ and very faintly in $\mu_{\phi_1}^*$.  }
            \label{fig:gd1-data_selection}
        \end{figure*}

@@

        The modeling framework, detailed in the \autoref{sec:method} sections, is
        modular, allowing many model components to be combined, nested, added
        to the mixture, and linked via priors.
        The model we require for \stream{GD-1} is similar to 
        the model built for the mock stream in \autoref{sec:results_mock:model},
        but has additional components for \stream{GD-1}'s spur.
        In this section we will discuss how the total model is built, noting 
        similarities to the example mocks-stream model.

>>>>

% Change the section names
\renewcommand{\sectionautorefname}{\S\!}%
\renewcommand{\subsectionautorefname}{\S\!\!}%
\renewcommand{\subsubsectionautorefname}{\S\!}%
\renewcommand{\equationautorefname}{Eq.}
\renewcommand{\tableautorefname}{Table}

@@

    % \begin{quotation}

    %     A stellar stream is the association of stars tidally stripped from a
    %     common progenitor system that orbits in an underlying galactic
    %     gravitational field \citep{Johnston1998, HelmiWhite1999}. Stream
    %     progenitors are gravitationally bound systems, such as satellite dwarf
    %     galaxies or star clusters \citep{Odenkirchen+2001, Majewski+2003}. In
    %     the absence of external interactions these systems are stable, though
    %     still capable of mass loss driven by internal processes
    %     \citep{Hills1975, HeggieHut2003}, e.g. relaxation and binary
    %     interactions. Considering external interactions, these progenitor
    %     systems may be disrupted: completely, as in the merger events that
    %     formed the Galactic stellar halo \citep{Lynden-Bell1967,
    %     SearleZinn1978}; or partially, such as tidal disruption forming a
    %     stellar stream.

    %     In the approximation of a smooth galactic gravitational field, the
    %     progenitor's extent is limited to the tidal radius. Through processes
    %     internal and external, progenitor stars can achieve sufficient energy to
    %     bring them beyond the tidal boundary, thus ``escaping'' into the
    %     galactic field. Escape primarily occurs at both an inner and outer
    %     saddle point of the combined potential -- e.g. the L2 and L3 Lagrange
    %     points for a spherical galactic field \citep{Ross+1997,
    %     FukushigeHeggie2000, BinneyTremaine2008}. Generally, the initial phase
    %     space position of the escaped star is very similar to that of the
    %     progenitor: the configuration position is the tidal boundary and the
    %     star has a low peculiar velocity. Since both the escaped star and its
    %     progenitor orbit in the same underlying galactic potential and have
    %     similar phase space positions their orbits are similar
    %     \citep{BinneyTremaine2008}. Given sufficient time the phase space
    %     separation will increase and an escaped star will lead or trail the
    %     position of the progenitor, depending on the location from which it
    %     escaped. Therefore, a collection of tidally stripped stars will form
    %     extended structures, aka tails or streams, that are few to thousands of
    %     parsecs in length \citep{Johnston1998, HelmiWhite1999, Bovy2014}.

    %     To be a coherent, observable structure the peculiar velocity dispersion
    %     of escaped stars relative to the progenitor must be relatively low, but
    %     does vary from system to system. Stars in kinematically hot streams
    %     (e.g. from progenitors with larger kinematic dispersions) have more
    %     dissimilar orbits from each other and the progenitor than do stars in
    %     cold streams, with smaller dispersions \citep{Johnston1998,
    %     HendelJohnston2015}.  The orbit of a star depends on its phase space
    %     position and the gravitational potential; therefore measuring a star's
    %     position and velocity and observing its orbit reveals the local
    %     potential, which is normally dominated by the Galaxy (primarily the dark
    %     matter halo at large distances) \citep{Gibbons+2014}. However, an orbit
    %     is impractical to observe on human timescales. Cold streams offer a
    %     workaround.  Like measuring one star at many points in one orbit over an
    %     extended period, streams are many stars measured at different points
    %     along approximately one orbit at the same time. Therefore stellar
    %     streams are one of the most promising means to study the Galactic
    %     gravitational potential: the stellar components as well as the dark
    %     matter \citep{Johnston2016}.

    %     On the observational front there has been a large concerted effort in
    %     the field to find new stellar streams \citep[for an extensive list of
    %     known Galactic streams see][]{Mateu2022}, extend the detected extents of
    %     known streams \citep[e.g. for \stream{Palomar 5} alone:][]{Rockosi+2002,
    %     Ibata+2017, Grillmair+2006, Carlberg+2012, Starkman+2019}, and increase
    %     the general purity of stream catalogs.

    % \end{quotation}

    % \todo[inline]{--- Rephrase}

    % \paragraph{one orbit and a single population}

    %     Stellar streams consist of stars that are tidally disrupted from a
    %     progenitor system.  To a first approximation, the stars occupy a single
    %     orbit, that of their progenitor.  Additionally, as stars formed in the
    %     progenitor they retain its photometric and chemical characteristics:
    %     which for globular clusters means following an isochrone.  In this
    %     simplistic picture, astrometrically characterizing a stream requires
    %     only finding it's orbit. Member stars fall directly on the orbit and the
    %     isochrone, making membership likelihood determination trivial.

    %     % Observational errors complicate this picture. Instead of a certain
    %     % membership calculation, we must build a probabilistic model, assigning
    %     % a likelihood that an observed star falls along the stream's orbit.

    % \paragraph{similar orbits and a somewhat noisy population}

    %     Occupying a single orbit implies that streams remain a point in action
    %     space, and having a common origin would thus not become extended
    %     structure. This is not true.  Differences in stream stars' orbits from
    %     the progenitor's arise from various factors including internal dynamics
    %     of the progenitor and the host Galactic potential.

    %     An intrinsic width, in configuration and kinematic space.

    %     Centered on a track, not an orbit.

    %     This defines an astrometric phase-space hyper-tube \cite{STREAMFINDER},
    %     with a distribution radial to the track.

    %     Instead of characterizing the stream by its orbit, we characterize it by
    %     it's track and the distribution around that track.  Computing membership
    %     likelihoods now requires fitting the track and distribution. This cannot
    %     be done without then also fitting the field of stars in which the stream
    %     lives, since the stream and field cannot be clearly separated.

    % \paragraph{gaps orbits and a somewhat noisy population}

    %     This is still a simplistic picture of streams.

    %     Perturbations to the Galactic potential, in the form of time evolution
    %     and non-axisymmetry, as well as substructure can disrupt the stellar
    %     stream.  Even most perturbations don't ruin the idea of the phase-space
    %     hypertube. They either drop the mixture weight (rephrase, not yet
    %     introduced) to 0 or change the width of the hypertube.

    % \paragraph{Review of detection methods}

    %     Identifying streams of globular of clusters is a challenging task.
    %     Streams are very low signal-to-noise structures spread over
    %     comparatively large regions of the sky. They live in a vast background
    %     of field stars.

    %     Historically streams were detected by matched-filter techniques
    %     % Add citations
    %     You took an isochrone and shifted it by a distance modulus until a
    %     stream stood out in longitude and latitude as an extended overdensity
    %     structure.

    %     With the advent of Gaia, we gained access to detailed astrometric data.
    %     For most streams this

    %     poses a challenging task due to their intricate and diffuse nature from
    %     large-scale surveys is challenging due to the complex and overlapping
    %     distributions of stars in different regions of the sky; disentangling
    %     members from the vast background of field stars

    %     \begin{enumerate}
    %         \item Streamfinder \citep{STREAMFINDER}. This method requires the
    %         potential
    %         \item VIA MACHINAE \citep{Shih+2022}
    %         \item Kiyan and Adrian
    %         \item Matched Filter
    %         \item Gaia stuff: Adrian and Ana and Sara et al
    %         \item Uniform Modeling of 12 Stellar stream \citep{Patrick+2022}.
    %         Only on phi1, phi2, and CMD. Their CMD method is more sophisticated.
    %         Worth discussing.
    %     \end{enumerate}

    % Review of available data:
    % \begin{enumerate}
    %     \item Gaia. Streams in Gaia have on-sky positions and proper motions,
    %     and parallaxes, though for most streams the parallax errors are
    %     prohibitively large. Likewise, most streams in Gaia do not have good
    %     radial velocity measurements. The photometry is generally poor,
    %     generally necessitating cross-matching with other datasets.
    %     \item S5. This is a much smaller dataset. It has deeper photometry and
    %     radial velocity measurements.
    % \end{enumerate}

@@

        Per \autoref{eq:stream_and_bkg_prob} the stream is an independent
        addition of both astrometric and photometric models.  In astrometrics
        the stream is a Gaussian track in $\phi_2$ and parallax $\parallax$, using
        the model defined in \autoref{ssub:method:astrometric_model:on_stream}.
        We use a 4-layer \texttt{LinTanh} with shape (1 input, 64 hidden, 4
        output), where the 4 outputs are $[\mu_{\phi_2}, \sigma_{\phi_2},
        \mu_{\parallax}, \sigma_{\parallax}](\phi_1)$.  The network is small, but
        sufficiently large to permit 15\% dropout and retain robust predictions.
        The photometric model is a distance modulus-shifted isochrone, using the
        model outlined in  \autoref{ssub:method:photometric_model:on_stream}. We
        use the same isochrone parameters as were used to simulate the stream.
        For this mock stream we demonstrate simultaneous astrometric and
        photometric modeling, linking the distance modulus and parallax track,
        as explained in \autoref{ssub:method:linking_to_astrometrics}. Thus the
        photometric model does not have its own network, using instead the
        outputs $[\mu_{\parallax}, \sigma_{\parallax}](\phi_1)$.

        The total background model has two pieces, an astrometric and a
        photometric model.  As a pre-training phase the photometric normalizing
        flow model was fitted to an off-stream selection and is now held fixed.
        The astrometric model is split in two components, the first is a uniform
        distribution in $\phi_2$, and the second is an exponential distribution
        for $\parallax$.  The uniform distribution does not have any parameters,
        and thus no neural network.  The exponential distribution has only the
        slope parameter and we use a 3-layer \texttt{LinTanh} with shape (1, 32,
        1) and 15\% dropout. Background and stream are combined in a mixture
        model whose mixture parameters are a 4-layer (1, 64, 1) \texttt{LinTanh}
        network. We train the mixture model with an \texttt{AdamW} optimizer
        with learning rate of $5\times10^{-3}$ and a scheduler to periodically
        and temporarily increase the learning rate, encouraging the optimizer
        towards better minima.

@@

        \begin{figure*}
            \script{mock/plot/results.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/mock/results.pdf}
            \caption{
                %
                \textbf{Top panel: }%
                    The stream mixture coefficient $f(\phi_1)$ (black dotted
                    line) predicted by the model, overplotted on the true stream
                    fraction (black-outlined histogram), calculated as the ratio
                    of stream stars to total stars in 55 bins over the $\phi_1$
                    range of the isochrone.  \textbf{Middle panel: }%
                    $\phi_2(\phi_1)$ over the full range of $\phi_1$. The data
                    color and transparency is set by the model's MLE stream
                    probability (with dropout turned off). Plotted behind the
                    data are the true labels for the stream stars (large black
                    circles).  The true labels exactly limn the model-identified
                    stream stars.  The MLE stream track (black dashed line) is
                    over-plotted on the probability-colored data.
                %
                \textbf{Lower panel 3: }%
                    $\parallax_2(\phi_1)$ over the full range of $\phi_1$. Like for
                    $\phi_2$, the data color and transparency is set by the MLE
                    probability, with the true labels (large black circles)
                    under-plotted and MLE track (black dashed line)
                    over-plotted.
                %
                \textbf{Upper Histograms: }%
                    $\phi_2$ histograms for each model component (blue
                    background, red stream), in 4 $\phi_1$ ranges across the
                    data set.  The height of each bin is the
                    probability-weighted density.  The background is a uniform
                    distribution, while the stream is a Gaussian. In projection
                    across the $\phi_1$ range it can appear multi-modal (panel
                    2) if there are large gaps or `smeared' out (panel 3) if the
                    Gaussian's mean changes significantly over the range.
                %
                \textbf{Upper Middle Histograms: }%
                    Parallax $\parallax$ histograms for each model component (blue
                    background, red stream), in 4 $\phi_1$ ranges across the
                    data set.  The height of each bin is the
                    probability-weighted density.
                %
                \textbf{Lower Middle and Lower Histograms:}%
                    Photometric coordinate $(g, r)$ plots for each model
                    component in 4 $\phi_1$ ranges across the data set. The top
                    row plots the stream stars (red) over the background. The
                    lower row plots the background stars (blue) over the stream
                    stars to show that the how the two distributions mix.
                %
            }
            \label{fig:mock_data_result}
            TODO:
            \begin{enumerate}
                \item the linear density in the stream-fraction plot
            \end{enumerate}
        \end{figure*}

@@

    Having successful applied the density modeling framework to a mock stream,
    we now apply the method to the stellar stream \stream{GD-1}.
    % TODO: some reiteration on the importance of GD-1
    We discuss the data selection in \autoref{sub:results_gd1:data}, model
    specifics in \autoref{sub:results_gd1:model}, and the results in
    \autoref{sub:results_gd1:results}.

@@

        We utilize data from \Gaia{} DR3 for the astrometric coordinates of each
        star.  To obtain more accurate color information, we cross-match the
        \Gaia{} field on \PanStarrs{} \citep{PanSTARRS1} using \Gaia's provided
        \texttt{Best-neighbors} catalog.  To limit the field to a region known
        to contain \stream{GD-1} we adopt broad cuts on the cross-matched data.
        We perform these rough cuts by requiring the data lie within the
        phase-space cube:
        \begin{itemize}
            \item $\phi_1 \in [-100, 40] \ \rm{\degree}$,
            \item $\phi_2 \in [-9, 5] \rm{\degree}$,
            \item $\parallax \in [-10, 1] \ \rm{mas}$,
            \item $\rm{bp-rp} \in [-1, 3] \ \rm{mag}$
            \item $g, r \in [0, 50] \ \rm{mag}$, and not \texttt{null}.
        \end{itemize}
        where we use \package{gala}'s \citep{gala, galav1.3}
        \code{GD1Koposov10} frame \citep{Koposov+2010} to define $\phi_1$,
        $\phi_2$, and associated proper motions.  For downloading purposes we
        split the query into 10 sub-regions in even $\phi_1$ increments. A plot
        of the $\phi_1, \phi_2$ regions is included in the
        \href{\GitHubURL{}}{\package{ShowYourWork}} repository.  Stream-specific
        coordinates ($\phi_1, \phi_2$) are not included in \Gaia.  One could
        embed the coordinate transform in the ADQL \citep{ADQL2.0}, or define
        great-circle arcs in ICRS \citep{ICRS1997}. Since only rough cuts are
        required and since the sub-region boxes are sufficiently small for a
        reasonable flat-sky approximation  we simply translate the ($\phi_1,
        \phi_2$) boxes to ICRS.  We correct the \PanStarrs{} photometry for dust
        extinction using the \textit{Bayestar19} \citep{Green+2019} models in
        the \package{dustmaps} package \citep{Green2018}.  \textit{Bayestar19}
        is a three-dimensional dust map of the Milky Way for which we use
        \Gaia's parallax distance information. For stars with missing parallax measurements we assume a distance of $8.5 \ \unit[kpc]$.  We include the $1-\sigma$ errors
        as a source of uncertainty in $\mbs{\Sigma}_n$, the per-star photometric
        uncertainty.
        We also correct the parallax for the zero-point using the \code{gaiadr3-zeropoint} package, which implements the findings from \cite{Lindegren+2021}.
        The result is a field populated by
        \variable{output/gd1/ndata_variable.txt} stars, from which we would like
        to identify stream members.

        We create data masks for a more refined selection around \stream{GD-1}.
        The two mask types, proper motion and photometric, are show in
        \autoref{fig:gd1-data_selection}.  The proper motion masks select
        rectangular regions, containing the primary overdensity of \stream{GD-1}
        and a surrounding area following the tail of the distribution.  The
        tighter mask (red) has corners $(\mu_{\phi_1}, \mu_{\phi_2}) =
        [(-15,-10) (-4.5, -2)] \ \unit{mas/yr}$, while the larger mask (yellow)
        has $(\mu_{\phi_1}, \mu_{\phi_2}) = [(-15,-5) (-5, -1)] \
        \unit{mas/yr}$.  In photometric coordinates we define a mask around
        % TODO: make these variables, like in mock
        % TODO: change the actual isochrone used in the code to -1.2, not -1.3
        a 12 Gyr, [Fe/H] = -1.2 isochrone at a distance of 7.8 kpc.
        The selection region is a 0.2 dex orthogonal buffer around the MIST
        isochrone \citep[using][]{brutus}, which is illustrated in the
        2nd-from-the-top-left plot in \autoref{fig:gd1-data_selection}.  In
        addition, to avoid detailed modeling of the completeness as a term in
        $\prior_{obs}$ (from \autoref{eq:isochrone_prior_components}) we mask out
        photometry with $G_{\Gaia{}} > 20$.  With combinations of these masks
        the stream has much larger signal-to-noise, though is still only a small
        percent of the total data.

        \begin{figure*}[h]
            \script{gd1/plot/data_selection.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/gd1/data_selection.pdf}
            \caption{%
                Data selections for \stream{GD-1}, within the data cube
                described in \autoref{sub:results_gd1:data}.
                %
                \textbf{Row 1 Panels:}
                Proper motion selection boxes in red and yellow.  The right plot
                shows the application of these selections.  \stream{GD-1} is
                apparent from around $-60$ to $-10$ deg in the red selection and
                present, if less clearly, in the yellow selection.
                %
                \textbf{Row 2 Panels:}
                Photometric selection (blue curve). The selection is based on a
                12 Gyr, [Fe/H] = -1.2 MIST isochrone, buffered by 0.2 dex.  The
                right plot shows the application of this selection. By eye, the
                stream is a small overdensity extending from $-50$ to $0$ deg.
                %
                \textbf{Row 3 Panels:}
                Applying the combination of red astrometric and blue photometric
                selections, plotted for astrometric coordinates $\phi_2, \parallax,
                \mu_{\phi_1}, \mu_{\phi_2}$.  The stream is an identifiable
                overdensity in all phase-space coordinates except $\parallax$ where
                the errors are significant.
                %
                \textbf{Row 4 Panels:}
                Applying the yellow astrometric and blue photometric selections,
                plotted for the same astrometric coordinates as the above row.
                In these coordinates the stream is clearly observable in
                $\phi_2$ and very faintly in $\mu_{\phi_1}^*$.  }
            \label{fig:gd1-data_selection}
        \end{figure*}

@@

        The modeling framework, detailed in the \autoref{sec:method} sections, is
        modular, allowing many model components to be combined, nested, added
        to the mixture, and linked via priors.
        The model we require for \stream{GD-1} is similar to 
        the model built for the mock stream in \autoref{sub:results_mock:model},
        but has additional components for \stream{GD-1}'s spur.
        In this section we will discuss how the total model is built, noting 
        similarities to the example mocks-stream model.

        All full Bayesian stream models necessarily include a background model,
        which we model in both astrometric and photometric spaces.
        The photometric background is too complex to capture with simple analytic
        distributions. Instead, we train a normalizing flow on the magnitudes $g, r$,
        conditioned on $\phi_1$. We refer readers to
        \autoref{sub:method:photometric_model:off_stream} for discussion of the photometric distribution and how a normalizing flow
        can characterize this complex distribution, including its variation over $\phi_1$.
        The flow's flexibility, while necessary in this context, also poses
        a problem: the flow captures the stream as easily as the background.
        Consequently we select on-stream and off-stream regions using the known
        $\phi_1,\phi_2$ range of the stream, as shown in the
        upper panel of \autoref{fig:gd1-photometric_background_selection}.
        In the lower panel are plotted these selections in photometric coordinates,
        where an isochrone is apparent in the on-stream selection.
        The normalizing flow is trained on the off-stream selection, and will be
        fixed (not trainable, nor contribute to the gradient) when incorporated into the larger model.
